Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-07-20 manuscript.tex
ongoing fixes

One of the most relevant difference between the SBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information \textcolor{red}{(\emph{i.e.} maximal entropy)} can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}, Fig. \ref{FigureRandom}A) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups, whereas nSBM does not identify any cell group \textcolor{red}{at default resolution} (Fig. \ref{FigureRandom}B).

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.pdf}
\caption[]{Modeling randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches (Adjusted Rand Index $ARI=0.869$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case nSBM does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
\end{figure}

\textcolor{red}{Only by reducing resolution to $\gamma \leq 0.5$ we were able to obtain a single partition by modularity (Fig. \ref{Modularity_Random_Res}). Of course,} this experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, it is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset. \textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modeled after the feature-wise distribution of detected genes. We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}). Of note, the concordance with original annotations drops at $\sigma \geq 1.5$ for both the approaches.}

@@

To benchmark \texttt{schist}, we tested it on scRNA-seq mixology data \cite{Tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data. In particular, we used the mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a considerable degree of heterogeneity and appears to be split into two cell groups (Fig. \ref{FigureTian}A). Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.829$), with a further split in H2228 cell line (Fig. \ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.549$, Fig. \ref{FigureTian}C). The nSBM solution correctly identifies H1975 groups as a single entity at level 1 of the hierarchy. We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes. Given the ground truth, the cell lines, SCCAF is able to assess the maximal accuracy that can be achieved in the dataset (0.992). When trained with this target accuracy, SCCAF precisely reconstructs the original cell line annotations starting from \texttt{schist} partitions with high accuracy (Fig. \ref{FigureTian}D). When Leiden partitions are set as input, SCCAF merges H2228 and HCC827 cells into a single cluster and keeps H1975 cells split into two groups (Fig. \ref{FigureTian}E), highlighting potential limitations of this approach. %In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig. \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig. \ref{Tian_NSBM0}B).

@@

When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \textcolor{red}{\emph{i.e.}} communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

\subsection*{Analysis of Random data}

Data were retrieved in \texttt{scanpy} environment using \texttt{sc.datasets.pbmc3k\_processed()} function. The random \emph{k}NN graph was obtained shuffling the node labels of each edge. UMAP embedding was recomputed after randomisation using the shuffled graph.

>>>>

One of the most relevant difference between the SBM and other methods to cluster single cells is that it relies on robust statistical modelling. In this sense, the number of groups identified strictly mirrors the amount of information contained in the data. An important consequence is that absence of information \textcolor{red}{(\emph{i.e.} maximal entropy)} can be properly handled. To show this property we performed a simple experiment on a randomised \emph{k}NN graph. We collected data for 3k PBMC (available as preprocessed data in \texttt{scanpy}, Fig. \ref{FigureRandom}A) and shuffled the edges of the prebuilt \emph{k}NN graph, this to keep the general graph properties unchanged. We tested that the degree distribution does not change after randomisation (Kolmogorov-Smirnov $D=0.0733$, $p=0.703$). We found that the default strategy, based on maximisation of modularity, identifies 24 cell groups \textcolor{red}{at default resolution}, whereas \textcolor{red}{\texttt{schist}} does not identify any cell group, \textcolor{red}{at level 0} (Fig. \ref{FigureRandom}B).

\begin{figure}[H]
\centering
\includegraphics[keepaspectratio,width=0.6\textwidth,height=\textheight]{FIgure_Random.pdf}
\caption[]{Modeling randomised data. (A) UMAP embedding of ~3k PBMC after standard processing with Leiden approach (left) or nSBM (right). Cell grouping is consistent for both the approaches (Adjusted Rand Index $ARI=0.869$). (B) UMAP embedding of the same data after randomisation of \emph{k}NN graph edges. In this case \texttt{schist} does not return any cell grouping, while optimisation of modularity finds up to 24 different cell groups}\label{FigureRandom}
\end{figure}

\textcolor{red}{Only by reducing resolution to $\gamma < 0.6$ we were able to obtain a single partition by modularity (Fig. \ref{Modularity_Random_Res}). Of course,} this experiment is a deliberate extreme case. The quality of grouping proposed by a standard approach can be disputed in many ways, and the UMAP embedding indeed reflects the absence of any information. Nevertheless, real-world data may include an unknown amount of random noise. Hence, it is important to identify cell groups that are not artefacts arising from processing and that do reflect the information contained in the dataset. \textcolor{red}{To understand the impact of random noise on structured data, we considered the same PBMC dataset and added white noise to the normalised counts at increasing levels of $\sigma$, ensuring that the noise level is modelled after the feature-wise distribution of detected genes. We then compared partitions to the original annotation by Adjusted Rand Index ($ARI$), we found that \texttt{schist} is more robust to perturbations and that, again, optimising the modularity results in overestimation of the number of communities at high noise levels (Table \ref{Table_Noise}). Of note, the concordance with original annotations drops at $\sigma \geq 1.5$ for both the approaches.}

@@

To benchmark \texttt{schist}, we tested it on scRNA-seq mixology data \cite{Tian_2019}, a dataset explicitly developed to benchmark single cell analysis tools without the need to simulate data. In particular, we used the mixture of 5 cell lines profiled with Chromium 10x platform. At a first evaluation of the UMAP embedding, all lines appear well separated. Only the lung cancer line H1975 shows a considerable degree of heterogeneity and appears to be split into two cell groups (Fig. \ref{FigureTian}A). Using default parameters, \texttt{schist} is able to identify correct cell groups ($ARI=0.829$), with a further split in H2228 cell line (Fig. \ref{FigureTian}B), whereas Leiden method clusters the dataset into 10 groups ($ARI=0.549$, Fig. \ref{FigureTian}C). \texttt{schist} correctly identifies H1975 groups as a single entity at level 1 of the nSBM hierarchy. We then sought to check if an independent agglomerative method, SCCAF \cite{miao_2020}, was able to recover cell line groupings starting from both partition schemes. Given the ground truth, the cell lines, SCCAF is able to assess the maximal accuracy that can be achieved in the dataset (0.992). When trained with this target accuracy, SCCAF precisely reconstructs the original cell line annotations starting from \texttt{schist} partitions with high accuracy (Fig. \ref{FigureTian}D). When Leiden partitions are set as input, SCCAF merges H2228 and HCC827 cells into a single cluster and keeps H1975 cells split into two groups (Fig. \ref{FigureTian}E), highlighting potential limitations of this approach. %In order to perform a fair and complete experiment, we also tested SCCAF with the partitions scheme at the deepest level of the nSBM hierarchy (38 groups, Fig. \ref{Tian_NSBM0}A), for which we observed the inability to perform a significant cell grouping (Fig. \ref{Tian_NSBM0}B).

@@

When grouping is performed by optimisation of modularity, there is often the implicit assumption that the resolution parameter reflects a hierarchical structure of the graph, \textcolor{red}{\emph{i.e.}} communities are consistently grouped at lower resolutions. Not only this assumption is wrong, but it may also lead to spurious groupings in real experiments, whereas a nSBM inherently encodes hierarchies by merging communities in a tree. The improper use of resolution parameter may lead to two types of errors: grouping of cells that are in fact distinct and creating an inconsistent hierarchy.

@@

\subsection*{Analysis of Randomized data}

Data were retrieved in \texttt{scanpy} environment using \texttt{sc.datasets.pbmc3k\_processed()} function. The random \emph{k}NN graph was obtained shuffling the node labels of each edge. UMAP embedding was recomputed after randomisation using the shuffled graph.
\textcolor{red}{To generate data with white noise we computed the genewise means ($\mu_g$) and standard deviations ($\sigma_g$) of log-normalized counts excluding 0 values. We generated random values using $\mu_g$ and $k\sigma_g$, $k \in \{0.5, 1, 1.5, 2\}$, and added to original expression values. }
